---
title: "The New Hospital Programme demand model"
subtitle: "A Tool for Forecasting Hospital Activity"
title-block-banner: true
output: html_document
execute:
  echo: false
  message: false
  warning: false
---

## What is it for?

The New Hospital Programme (NHP) is building hospitals across England. To help plan for the future, [The Strategy Unit](https://www.strategyunitwm.nhs.uk/) developed a tool to predict how many people might use these hospitals and how long they’ll need care. 

## What does it do?

The model estimates:

- How many people will need treatment (admissions).
- How long patients might stay in the hospital (bed days)

It looks at all types of hospital visits, like surgeries, check-ups, and emergencies, and breaks it down by things like age, sex, and treatment.

```{r multiple}
value_boxes <- list(
  bslib::value_box(
    title = "NHS trusts using the model",
    value = "27+",
    showcase = bsicons::bs_icon("hospital"),
    theme = "primary"
  ),
  bslib::value_box(
    title = "Looking ahead",
    value = "2041/42",
    showcase = bsicons::bs_icon("graph-up"),
    theme = "primary"
  ),
  bslib::value_box(
    title = "Customisation options",
    value = "90+ ",
    showcase = bsicons::bs_icon("gear"),
    theme = "primary"
  ),
  bslib::value_box(
    title = "Rows of real hospital data",
    value = "140 million",
    showcase = bsicons::bs_icon("journal-medical"),
    theme = "primary"
  )
)


bslib::layout_column_wrap(
  width = "150px",
  !!!value_boxes[1:2]
)

```

## What factors does it consider?

The tool considers many things that can change how much people need hospitals in the future:

### Population changes

As the population grows, ages, and experiences changes in health, healthcare demand will change. We use ONS population projections and health trends to model how these factors will evolve over time.

### Non-demographic growth

This refers to changes in healthcare demand not related to population size, age, or gender, such as:

- Advances in medical technology
- Changes in clinical standards
- Rising patient expectations

Our experts have analysed previous trends to predict how this growth will affect the future.

### Prevention 

Hospitals use strategies to reduce healthcare activity, such as preventing falls in older people or shortening hospital stays. These strategies include:

- Shifting care to local settings
- Preventing health issues early
- Improving efficiency

NHS trusts tell us which of these strategies they intend to implement, and estimate how much they might reduce future activity. They provide these as a _range_, rather than a single numbner, for example a reduction of between 10% and 30% of elderly fall treatments.

```{r}
bslib::layout_column_wrap(
  width = "150px",
  !!!value_boxes[3:4]
)
```


## Many possible futures

The future is uncertain. We run our model hundreds of times to generate a range of possible futures, based on the data trusts provide. 

```{r, eval=TRUE}
library(ggplot2)

# Parameters
n_rep <- 10
baseline <- 500
targets <- 500 + rnorm(n = n_rep, mean = 120, sd = 50)
targets <- c(targets, mean(targets))
steps <- 22               
n_runs <- length(targets) 

random_walk <- function(start, end, steps) {
  increments <- (end - start) / steps + rnorm(steps, mean = 0, sd = abs(end - start) / (2 * steps))
  cumsum(c(start, increments)) 
}

# Generate data for all runs
set.seed(42) 
random_walks <- tibble::tibble(
  run = seq_along(targets),
  target = targets
) |>
  dplyr::mutate(
    data = purrr::map(target, \(x) tibble::tibble(
      step = 0:steps,
      value = random_walk(baseline, x, steps)
    ))
  ) |>
  tidyr::unnest(data) 

random_walks |>
  dplyr::mutate(step = step + 2019) |>
  ggplot(aes(x = step, y = value)) + 
  geom_line(aes(group = run), colour = "#005EB8", linewidth = 1.8) + 
  gghighlight::gghighlight(run == n_runs, use_direct_label = FALSE) + 
      labs(title = "Our model provides hundreds of possible futures and the <span style=\'color:#005EB8'>**average forecast**</span>.") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none",
        plot.title= ggtext::element_markdown(),
        panel.grid = element_blank(),
        panel.border = element_blank()) 
```

::: {.callout-note}
## Learn more 

Explore additional resources:

- `r bsicons::bs_icon("youtube")` Watch a [video](https://www.youtube.com/watch?v=vXBxDOa6tfw)
- `r bsicons::bs_icon("book")` Explore the [documentation](https://connect.strategyunitwm.nhs.uk/nhp/project_information)
- `r bsicons::bs_icon("chat-dots")` [Talk]() to us
:::
